Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-06-25 Manuscript/manuscript.tex
Manuscript edits and sensitivity analysis
Edits throughout manuscript; incorporating sensitivity analysis into
results and discussion; adding sensitivity analysis results.

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.  In particular, we expect that the host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.  Indeed, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes in turn may then influence where environmental pathogen reservoirs get established and how they affect disease dynamics and host occupancy throughout the metapopulation.

@@

\subsection{Model Structure}
To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of four possible states: occupied by susceptibles (S), occupied by infecteds (I), pathogen reservoir without host population (Z), and unoccupied (by both pathogen and host) ($\oslash$).  State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infected either through direct contact with infected immigrants, or through a local pathogen reservoir.

@@

The extinction rate of patch $i$ in state $X$ is given by:
\begin{equation}
E_{Xi}=\frac{e_X}{A_i^\alpha},
\end{equation}
where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$).  When an infected patch goes extinct, we assume that the infected hosts leave behind an infectious pathogen reservoir.

Susceptible populations are infected via direct contact with infected colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.  Environmental transmission occurs when a susceptible population occupies a reservoir patch, and occurs at rate:
\begin{equation}
\gamma=\gamma_0exp(-rt_I),
\end{equation}
where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_I$ is the time since last infected occupancy, and $r$ is the pathogen's decay rate in the environment.

\begin{table}[h!]
\label{parameters}      
\caption{Parameters of SPOM model and their meanings.}
\begin{tabular}{l p{8.5cm} r}
Parameter & Interpretation &  Default Value\\
\hline
$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5\\
$\xi_{em}$ & Scaling parameter for effect of target patch quality in emigration & 0.5\\
$D$& Inverse of mean dispersal distance & 5\\
$d_{ij}$ & Distance between patch $i$ and $j$ & 1 for all $i,j$\\
\hline
$e_S$ & Extinction rate of unit quality susceptible patch & 0.1\\
$e_I$ & Extinction rate of unity quality infected patch & 0.5\\
$\alpha$ & Strength of environmental stochasticity & 1\\
\hline
$\delta$ & Probability of direct infection & 0.5\\
$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5
\end{tabular}
\end{table}

@@

The model outlined above was parameterized (Table 1) according to the following assumptions. (1) High quality patches maintain higher population sizes and therefore produce more colonists than low quality patches and have lower extinction rates than low quality patches ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  (2) High quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).  (3)  Connectivity is determined entirely by quality ($d_{ij}=1$).  In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($e_S = 0.1$, $D=5$), and a range of epidemiological behaviors were feasible ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).

With the above parameters fixed, we explored the effects of pathogen environmental persistence and habitat heterogeneity by varying pathogen longevity (10 values from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity), and the variance of the patch quality distribution (10 values ranging from 0.02 to 0.2).  For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with mean 1 and given variance. In each case, an entirely susceptible population was simulated until it reached a steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  To address the metapopulation-level questions above, we recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infected populations after 5000 time steps), endemic (both susceptible and infected populations present), or pandemic (only infected populations).  In addition, to address patch-level questions, for every simulation, we recorded the quality of each patch and the proportion of time spent in each of the four states.

In the above simulations, the mean of the patch quality distribution was held constant, and thus the range was expanded symmetrically about one as the variance was increased.  To further parse out the relative roles of high and low quality patches, we performed a series of simulations where we expanded the range of the quality distribution non-symmetrically to include high and low quality patches separately.  For these simulations, we fixed longevity at an intermediate level (100) and ran 100 replicates for each of four patch quality distributions: low varaince, (range 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).

@@

The pathogen persisted in 7087 of the 10000 simulations, spreading through the entire susceptible host population (i.e. a pandemic) in 3721 simulations.  Pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).  The effect of longevity, however, depended on the variance of the patch quality distribution, with low variance metapopulations favoring endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowering the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).

@@

To separate the influence of high and low quality patches, we explored how susceptible and infected patch occupancy changes when high and low quality patches are included separately in the patch quality distribution.  Relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution toward low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest effects observed at intermediate longevities (Fig~\ref{sens}(a)).  Conversely, a shift towards low quality patches decreased infected occupancy, while a shift towards high quality patches increased infected occupancy (Fig~\ref{sens}(b)).

However, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).  Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread infection.  On the other hand, at high pathogen longevities, the positive effect of low qualtiy distributions on susceptible occupancy is overwhelmed, and low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events.  Thus, the role of the patch quality distribution and the presence of high or low quality patches is generally greatest for low to intermediate pathogen longevity -- once longevity is high enough, the pathogen persists and spreads easily enough that the effects of individual patches are washed out.  Then the role of patch quality lies primarily in its effect on extinction rates, and whether infected hosts can persist.

@@

To understand the mechanisms behind these effects, we further examined within patch dynamics in high variance metapopulations that produced endemic dynamics.  Because of their lower extinction rate and higher recolonization rate, high quality patches support more consistent occupancy than low quality patches (Fig~\ref{simvis}).  However, as an epidemic progresses, high quality patches increasingly support infected occupants, while low quality patches support infecteds infrequently (Fig~\ref{simvis}).  In addition, while high quality patches do support susceptible occupancy, they experience a greater number of infection events (susceptible to infected transitions) than low quality patches (Fig~\ref{infections}).  This relationship becomes weaker as longevity increases, due to the fact that as longevity increases, there are fewer susceptibles remaining in the metapopulation to become infected.

@@

The positive relationship between pathogen spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.  \cite{Almberg2011} find that, in plausible simulations of chronic wasting disease in deer, increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the  importance of reservoir transmission increasing through time.  Similarly, Breban et al 2009 show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.  \emph{Other examples.}

Moreover, for a given longevity, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}.  The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.  In this context, our metapopulation can be viewed a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).

These simulations also offer insight into the interaction between environmental longevity and habitat heterogeneity, suggesting that the specific effect of longevity on pathogen spread is mediated by the variance of the patch quality distribution.  In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).  In this way, the effects of environmental longevity are exacerbated, or amplified, by a high variance patch quality distribution, especially at intermediate longevities.

This interaction between environmental longevity and the variance of the patch quality distribution is largely driven by the presence of high quality patches.  Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.  These two properties have important implications for pathogen spread in the metapopulation. Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure results in part from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}), but is further augemented by environmental transmission.

This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}).  Once the pathogen infects high quality patches, the larger number of colonists produced allow it to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infected occupancy throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infected occupancy in the metapopulation (Fig~\ref{sens}).

Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by facilitating increased susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches are unable to support consistent occupancy (Fig~\ref{simvis}), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation, an effect that is strongest at intermediate longevities (Fig~\ref{sens}).   This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infected populations, means that infected populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infected occupancy (Fig~\ref{sens}).

However, the importance of the patch quality distribution in pathogen spread depends on the pathogen's longevity, with the strongest effects of patch quality observed for intermediate longevities.  For these longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.  In contrast, for low longevities, when the pathogen reservoir decays much faster than colonizations and extinctions occur, environmental transmission has little effect on disease dynamics and the effect of patch quality is relatively weak.  Conversely, when the pathogen decays much slower than colonizations and extinctions occur, i.e. high longevity, the pathogen reservoir persists on nearly all patches and the effects of patch quality are washed out.  In these cases, the role of patch quality lies primarily in its effect on extinction rates.  Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread (pandemic) infection, while low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events at the highest longevities.

@@

@@

>>>>

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.  In particular, we expect that host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes in turn may then influence where environmental pathogen reservoirs get established and how they affect disease dynamics and host occupancy throughout the metapopulation.

@@

\subsection{Model Structure}
To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infected hosts (I), and unoccupied by the host.  State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infected either through direct contact with infected immigrants, or through a local pathogen reservoir.

@@

The extinction rate of patch $i$ in state $X$ is given by:
\begin{equation}
E_{Xi}=\frac{e_X}{A_i^\alpha},
\end{equation}
where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$), and $\alpha$ controls how extinction rate scales with patch quality.  When an infected population goes extinct, we assume that the infected hosts leave behind an infectious pathogen reservoir on the patch.

Susceptible populations are infected via direct contact with infected colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.  In addition, environmental transmission can take place when a susceptible population occupies a patch with an environmental reservoir, and occurs at rate:
\begin{equation}
\gamma_i=\gamma_0exp(-rt_{Ii}),
\end{equation}
where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_{Ii}$ is the time since last infected occupancy, and $r$ is the pathogen's decay rate in the environment.

\begin{table}[h!]
\label{parameters}      
\caption{Parameters of SPOM model and their interpretations.}
\begin{tabular}{l p{8.5cm} r}
Parameter & Interpretation &  Default Value\\
\hline
$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5\\
$\xi_{em}$ & Scaling parameter for effect of target patch quality in emigration & 0.5\\
$D$& Inverse of mean dispersal distance & 5\\
$d_{ij}$ & Distance between patch $i$ and $j$ & 1 for all $i,j$\\
\hline
$e_S$ & Extinction rate of unit quality susceptible patch & 0.1\\
$e_I$ & Extinction rate of unity quality infected patch & 0.5\\
$\alpha$ & Strength of environmental stochasticity & 1\\
\hline
$\delta$ & Probability of direct infection & 0.5\\
$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5
\end{tabular}
\end{table}

@@

The model outlined above was parameterized (Table 1) according to the following assumptions. (1) High quality patches maintain higher population sizes and therefore produce more colonists than low quality patches and have lower extinction rates than low quality patches ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  (2) High quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).  (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($e_S = 0.1$, $D=5$), and a range of epidemiological behaviors were feasible ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).

With the above parameters fixed, we explored the effects of pathogen environmental persistence and habitat heterogeneity by varying pathogen longevity (10 values ranging from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity), and the variance of the patch quality distribution (10 values ranging from 0.02 to 0.2).  For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with mean 1 and given variance. In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  To address the metapopulation-level questions above, we recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infected populations after 5000 time steps), endemic (both susceptible and infected populations present), or pandemic (only infected populations).  In addition, to address patch-level questions, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state, and the number of transitions between each state.

In the above simulations, the mean of the patch quality distribution was held constant, and thus the range was expanded symmetrically about one as the variance was increased.  To further parse out the relative roles of high and low quality patches, we performed a series of simulations where we expanded the range of the quality distribution non-symmetrically to include high and low quality patches separately.  For these simulations, at each value for longevity, we simulated 100 metapopulations for each of four patch quality distributions: low varaince, (range 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).

In addition, to explore the effects of the model assumptions and parameterization, we repeated the above analyses for three additional model parameterizations: with patches arranged in a lattice structure; with $\alpha = 0$ to remove the effect of patch quality on extinction rate; and with $\delta = 0.1$ to reduce the relative importance of direct transmission.

@@

The pathogen persisted in 7087 of the 10000 simulations, spreading through the entire susceptible host population (i.e. a pandemic) in 3721 simulations.  Pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).  The effect of longevity, however, depended on the variance of the patch quality distribution, with low variance metapopulations favoring endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowering the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).  The three alternative parameterizations produced qualitatively similar results, though pathogen spread was generally more difficult under these assumptions (i.e. fewer pandemics), and the habitat quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).

@@

To separate the influence of high and low quality patches, we explored how susceptible and infected patch occupancy changes when high and low quality patches are included separately in the patch quality distribution.  Relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution toward low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest effects observed at intermediate longevities (Fig~\ref{sens}(a)).  Conversely, a shift towards low quality patches decreased infected occupancy, while a shift towards high quality patches increased infected occupancy (Fig~\ref{sens}(b)).  These general trends were consistent across the three other parameterizations (Figs~\ref{sens_lattice}, \ref{sens_x0}, \ref{sens_delta}).

However, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).  Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high longevity.  On the other hand, when high pathogen longevities lead to pandemic dynamics, low quality distributions were unable to support infected occupancy, resulting in an increase in host extinction events and a decrease in mean occupancy.  The model in which the extinction rate does not scale with quality (i.e. $\alpha = 0$) produced very similar results (Fig~\ref{sens_x0}(c)).  However, in simulations with weak direct infection ($\delta = 0.1$), the high quality patch distributions always produced higher occupancy than low quality patch distributions (Fig~\ref{sens_delta}(c)), while in the model with the lattice structure, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevity (Fig~\ref{sens_lattice}(c)).

@@

To understand the mechanisms behind these effects, we further examined within patch dynamics in high variance metapopulations (with qualities ranging from 0.25 to 1.75) that produced endemic dynamics.  Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  However, as the epidemic progressed, high quality patches increasingly supported infected occupants, while low quality patches supported infecteds only briefly and infrequently (Fig~\ref{simvis}).  In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infected transitions) than low quality patches (Fig~\ref{infections}).  This relationship became weaker as longevity increased, due to the fact that as longevity increases and the pathogen spreads more easily, there are fewer susceptibles remaining in the metapopulation to become infected.  A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}).

@@

The positive relationship between pathogen spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.  \cite{Almberg2011} find that, in plausible simulations of chronic wasting disease in deer, increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the  importance of reservoir transmission increasing through time.  Similarly, \cite{Breban2009} show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.  \emph{Other examples.}

Moreover, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}).  In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.  In this context, our metapopulation can be viewed a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).  However, as the alternative parameterizations show, the overall effect of patch quality variance depends on the structure of the metapopulation, the resulting colonization and extinction rates, and the relative balance of the effects of low and high quality patches.  In the focal model, the dynamics of high variance metapopulations were driven largely by the presence of high quality patches, whereas in the models with the lattice structure and with equal extinction rates across all patches, high and low quality patches were more balanced in their effects, thus canceling each other out as variance increased.

The effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches in pathogen spread and metapopulation stability.  Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.  These two properties have important implications for pathogen spread in the metapopulation. Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure results from both transmission from the environmental reservoir and from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  Once the pathogen infects high quality patches, the larger number of host colonists produced by high quality patches allow it to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infected occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infected colonists that help maintain the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infected occupancy in the metapopulation (Fig~\ref{sens}).

In contrast, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches are unable to support consistent occupancy (Fig~\ref{simvis}), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).   This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infected populations, means that infected populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infected occupancy (Fig~\ref{sens}).

However, the ability of low quality patches to influence disease dynamics depends on the relative strength of the different transmission pathways -- direct and environmental -- and on the structure of the metapopulation.  At intermediate longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.  In contrast, for low longevities, when the pathogen reservoir decays much faster than colonizations and extinctions occur, environmental transmission has little effect on disease dynamics and the effect of patch quality is relatively weak.  Conversely, when the pathogen decays much slower than colonizations and extinctions occur, i.e. high longevity, the pathogen reservoir persists on nearly all patches and the effects of patch quality are washed out.  In these cases, the role of patch quality lies primarily in its effect on extinction rates.  Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread (pandemic) infection, while low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events at the highest longevities.

@@

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{metapop(lattice).pdf}
\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome_lattice}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{highlow(lattice).pdf}
\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.  Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
\label{sens_lattice}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{infevents(lattice).pdf}
\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.  The thin black lines show the smoothed fits for longevities of 40 (bottom) and 140 (top).}
\label{infections_lattice}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{metapop(delta).pdf}
\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
\label{poutcome_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{highlow(delta).pdf}
\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.  Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
\label{sens_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{infevents(delta).pdf}
\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.  The thin black lines show the smoothed fits for longevities of 60 (bottom) and 120 (top).}
\label{infections_delta}
\end{figure}
